Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-22 paper.tex
First version of OFH narrative

\subsection{Case study: Our Future Health genotype release}
Our Future Health~\cite{ofhmain_2025} is the UK’s largest ever health research programme. 
It is designed to help people live healthier lives for longer through the 
discovery and testing of more effective approaches to prevention, earlier 
detection and treatment of diseases.
Our Future Health is inviting millions of people, from all backgrounds and 
from right across the UK to take part. Volunteers are providing information 
about their health and lifestyles to create an incredibly detailed picture 
that represents the whole of the UK.

% this section is probably better placed in acknowledgements?
This study makes use of de-identified data held by Our Future Health. 
We would like to acknowledge all the research participants who have donated 
their data to the Our Future Health research programme 
\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}

Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
was made available in December 2024. The genetic component includes genotype data 
from 651,050 individuals on 707,522 variants. Here we use the genotype data from 
chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
10,221 variants.

@@

>>>>

\subsection{Case study: Our Future Health genotype data}
Our Future Health~\cite{ofhmain_2025}  is the UK’s largest ever health research programme. 
It is designed to help people live healthier lives for longer through the 
discovery and testing of more effective approaches to prevention, earlier 
detection and treatment of diseases.
Our Future Health is inviting millions of people, from all backgrounds and 
from right across the UK to take part. Volunteers are providing information 
about their health and lifestyles to create an incredibly detailed picture 
that represents the whole of the UK.

\begin{table}
\caption{Summary of the largest VCF Zarr columns for the Our Future Health
    genotype dataset, with chunk sizes of 1,000 variants and 
    10,000 samples and applying the BitRound(5) filter to the floating point 
    fields.
    For column details, see Table~\ref{tab-genomics-england-data}.
    \label{tab-ofh-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
/call\_LRR & float32 & 8.41 GiB & 2.9 & 51.65\% \\
/call\_BAF & float32 & 6.14 GiB & 4.0 & 37.71\% \\
/call\_GS & float32 & 1021.24 MiB & 25.0 & 6.12\% \\
/call\_genotype & int8 & 663.97 MiB & 19.0 & 3.98\% \\
/call\_genotype\_mask & bool & 84.7 MiB & 150.0 & 0.51\% \\
/sample\_id & object & 3.19 MiB & 1.6 & 0.02\% \\
/call\_genotype\_phased & bool & 1.08 MiB & 5900.0 & 0.01\% \\
/variant\_id & object & 40.83 KiB & 2.0 & 0.00\% \\
/variant\_position & int32 & 39.03 KiB & 1.0 & 0.00\% \\
/variant\_allele & object & 20.71 KiB & 7.7 & 0.00\% \\
/variant\_filter & bool & 12.55 KiB & 0.8 & 0.00\% \\
/variant\_quality & float32 & 4.51 KiB & 8.8 & 0.00\% \\
/variant\_contig & int8 & 4.51 KiB & 2.2 & 0.00\% \\
/contig\_id & object & 4.51 KiB & 0.0 & 0.00\% \\
/variant\_id\_mask & bool & 4.5 KiB & 2.2 & 0.00\% \\
/contig\_length & int64 & 4.5 KiB & 0.0 & 0.00\% \\
/filter\_id & object & 4.43 KiB & 0.0 & 0.00\% \\
\bottomrule
\bottomrule
\end{tabular}
\end{table}

Release 9~\citep{ofhdatarelease9}
was made available in December 2024. The genetic component includes genotype data 
from 651,050 individuals on 707,522 variants. Here we use the genotype data from 
chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
% 20G     ofh.chr22-b0.vcf.gz
% 19G     ofh.chr22-b1.vcf.gz
% 3.3G    ofh.chr22-b2.vcf.gz
% ==42.3 GiB - assuming the G is power-of-two here
10,221 variants and 42.3GiB of storage. We first converted the VCF 
data to ICF, which took 8m 55s elapsed (138m CPU) using 20 processes.
The ICF representation required a total of 41GiB over XX files.
We then converted to VCF Zarr format using the default schema, which resulted
in a 41GiB VCZ. On inspection, this was dominated by three floating point 
fields LRR, BAF and GS stored 
to a high degree of precision (7 decimal places for LRR and BAF and 4 decimal
places for GS).
On inspecting the histograms for these fields
(Fig~\ref{fig-ofh-field-distributions}) it seemed that such high precision
was likely unneccessary, and we investigated applying the Quantize
and Bitround filters from \texttt{numcodecs}~\citep{numcodecs}
to reduce the precision in stored floating point data, 
and therefore improve compression.
Using the Quantize filter with 5 digits of precision on these three fields reduced the 
overall storage space to 31G with a mean absolute error (MAE, i.e., the average
difference between the truncated and original values) of 
% 3   call_LRR    quantize    1.91e-06
% 4   call_BAF    quantize    1.18e-06
% 5   call_GS quantize    1.88e-06
$1.91\times10^{-6}$ (LRR), 
$1.18\times10^{-6}$ (BAF)  and 
$1.88\times10^{-6}$ (GS). 
Using the Bitround filter with a parameter of 5 bits reduced the overall storage for the Zarr to
XX with an MAE of
% 0   call_LRR    bitround    6.93e-04
% 1   call_BAF    bitround    7.86e-04
% 2   call_GS bitround    3.38e-03
$6.93\times10^{-4}$ (LRR), 
$7.86\times10^{-4}$ (BAF)  and 
$3.38\times10^{-3}$ (GS). 
The histograms of the truncated fields in all cases were visually indistinguishable 
from the original in Fig~\ref{fig-ofh-field-distributions} (data not shown).
Table~\ref{tab-ofh-data} shows a summary of the Zarr using the BitRound filter.

@@

\begin{figure}[h]
\begin{tabular}{ccc}
\includegraphics[width=2.5cm]{ofh_example/plots/default_call_LRR.png}&
\includegraphics[width=2.5cm]{ofh_example/plots/default_call_BAF.png}&
\includegraphics[width=2.5cm]{ofh_example/plots/default_call_GS.png}
\end{tabular}
\caption{Distribution of floating point fields in the Our Future Health data.
\label{fig-ofh-field-distributions}}
\end{figure}
